noam_beta_diversity

### 2024-09-06 noam_beta_diversity.Rmd
added beta dispersion for quad level

bray_mod_site <- glmmTMB(dissimilarity_trans ~ time*habitat + (1 | site),
                              na.action = na.pass, family = beta_family(link = "logit"), data = bray_time_series_site) #
summary(bray_mod_site) 
hist(residuals(bray_mod_site)) # slight skew but not awful
plot(residuals(bray_mod_site) ~ fitted(bray_mod_site)) # some heteroskedasticity but not bad

emtrends(bray_mod_site, pairwise ~ habitat , "time") # no sig differences
bray_trends_emm_site <- emmip(bray_mod_site, habitat ~ time , at = list(time = seq(1,16,1)), plotit = F, CIs = T)

@@

```{r dissimilarity compared to baseline all, fig.height=15, fig.width = 15}
bray_time_series_site_baseline <- subset(bray_time_series_site, year2 == 2007)
bray_mod_site_baseline <- glmmTMB(dissimilarity_trans ~ time*habitat + (1 | site),
                              na.action = na.pass, family = beta_family(link = "logit"), data = bray_time_series_site_baseline)
hist(residuals(bray_mod_site_baseline)) # long tails but not bad
plot(residuals(bray_mod_site_baseline) ~ fitted(bray_mod_site_baseline) ) # heteroskedastic

emtrends(bray_mod_site_baseline, pairwise ~ habitat , "time") # fringing slope significantly different from all others
bray_trends_emm_site_baseline <- emmip(bray_mod_site_baseline, habitat ~ time , at = list(time = seq(1,16,1)), plotit = F, CIs = T)

@@

```{r dissimilarity compared to baseline macro, fig.height=15, fig.width = 15}
bray_time_series_site_macro_baseline <- subset(bray_time_series_site_macro, year2 == 2007)
bray_mod_site_macro_baseline <- glmmTMB(dissimilarity_trans ~ time*habitat + (1 | site),
                              na.action = na.pass, family = beta_family(link = "logit"), data = bray_time_series_site_macro_baseline)
hist(residuals(bray_mod_site_macro_baseline)) # slight negative skew
plot(residuals(bray_mod_site_macro_baseline) ~ fitted(bray_mod_site_macro_baseline) ) # strong negative trend

emtrends(bray_mod_site_macro_baseline, pairwise ~ habitat , "time") # fringing slope significantly different from all others
bray_trends_emm_site_macro_baseline <- emmip(bray_mod_site_macro_baseline, habitat ~ time , at = list(time = seq(1,16,1)), plotit = F, CIs = T)

@@

full_columns_to_keep <- apply(full_site_species_matrix, 2, function(x) any(x >= asin(sqrt(0.1) )))
full_abundant_site_species_matrix <- as.data.frame(full_site_species_matrix)[ , full_columns_to_keep]
full_abundant_site_species_names <- colnames(full_abundant_site_species_matrix)

@@

# read in disturbances
disturbances <- read.csv(here("data", "disturbance_list.csv"))
full.cent_site <- merge(full.cent_site, disturbances, by = "year") # add disturbances to path
library(ggrepel)
ggplot() + 
  geom_point(data= full.site.scores_site, aes(x=NMDS1,y=NMDS2, colour = habitat), alpha = 0.05) + # add the point markers
  geom_text_repel(data= full.abundant.species.scores_site,aes(x=NMDS1,y=NMDS2, label = species), size = 4, alpha = 0.5) +  # add the species labels

  geom_path(data = full.cent_site, aes(x = NMDS1, y = NMDS2, group = habitat, colour = habitat),
            arrow = arrow(length=unit(0.25,"cm"), ends="last", type = "closed"), size = 1) +
  geom_point(data = full.cent_site, aes(x = NMDS1, y = NMDS2, group = habitat, colour = habitat, shape = disturbance), size = 3, stroke = 1.4) +
  scale_shape_manual(values = c(
    "cots" = 8,
    "cyclone" = 13,
    "bleaching" = 4
  )) +

@@

full_columns_to_keep <- apply(full_site_macro_species_matrix, 2, function(x) any(x >= asin(sqrt(0.1) )))
full_abundant_site_macro_species_matrix <- as.data.frame(full_site_macro_species_matrix)[ , full_columns_to_keep]
full_abundant_site_macro_species_names <- colnames(full_abundant_site_macro_species_matrix)

@@

----------------- QUADRAT BELOW
```{r fringing bray curtis}
fringing_species_matrix <- as.matrix(fringing[ , -which(names(fringing) %in% meta_cols)])
fringing_species_matrix <- fringing_species_matrix/100 # convert from percent to proportion
rownames(fringing_species_matrix) <- paste(fringing$year,"-", fringing$location, sep = "")

# filter rare species
fringing_species_presence <- colSums(fringing_species_matrix > 0) # Presence absence matrix
fringing_species_percentage <- (fringing_species_presence/nrow(fringing_species_matrix)) # proportion of rows where a species is present
fringing_common_species <- names(fringing_species_percentage[fringing_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
fringing_species_matrix <- fringing_species_matrix[,colnames(fringing_species_matrix) %in% fringing_common_species] # filter original species matrix

fringing_rows_to_remove <- rowSums(fringing_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
fringing_species_matrix <- fringing_species_matrix[!fringing_rows_to_remove,]

@@

```{r backreef bray curtis}
backreef_species_matrix <- as.matrix(backreef[ , -which(names(backreef) %in% meta_cols)])
backreef_species_matrix <- backreef_species_matrix/100 # convert from percent to proportion
rownames(backreef_species_matrix) <- paste(backreef$year,"-", backreef$location, sep = "")

# filter rare species
backreef_species_presence <- colSums(backreef_species_matrix > 0) # Presence absence matrix
backreef_species_percentage <- (backreef_species_presence/nrow(backreef_species_matrix)) # proportion of rows where a species is present
backreef_common_species <- names(backreef_species_percentage[backreef_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
backreef_species_matrix <- backreef_species_matrix[,colnames(backreef_species_matrix) %in% backreef_common_species] # filter original species matrix

backreef_rows_to_remove <- rowSums(backreef_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
backreef_species_matrix <- backreef_species_matrix[!backreef_rows_to_remove,]

@@

```{r outer10 bray curtis}
outer10_species_matrix <- as.matrix(outer10[ , -which(names(outer10) %in% meta_cols)])
outer10_species_matrix <- outer10_species_matrix/100 # convert from percent to proportion
rownames(outer10_species_matrix) <- paste(outer10$year,"-", outer10$location, sep = "")

# filter rare species
outer10_species_presence <- colSums(outer10_species_matrix > 0) # Presence absence matrix
outer10_species_percentage <- (outer10_species_presence/nrow(outer10_species_matrix)) # proportion of rows where a species is present
outer10_common_species <- names(outer10_species_percentage[outer10_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
outer10_species_matrix <- outer10_species_matrix[,colnames(outer10_species_matrix) %in% outer10_common_species] # filter original species matrix

outer10_rows_to_remove <- rowSums(outer10_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
outer10_species_matrix <- outer10_species_matrix[!outer10_rows_to_remove,]

@@

```{r outer17 bray curtis}
outer17_species_matrix <- as.matrix(outer17[ , -which(names(outer17) %in% meta_cols)])
outer17_species_matrix <- outer17_species_matrix/100 # convert from percent to proportion
rownames(outer17_species_matrix) <- paste(outer17$year,"-", outer17$location, sep = "")

# filter rare species
outer17_species_presence <- colSums(outer17_species_matrix > 0) # Presence absence matrix
outer17_species_percentage <- (outer17_species_presence/nrow(outer17_species_matrix)) # proportion of rows where a species is present
outer17_common_species <- names(outer17_species_percentage[outer17_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
outer17_species_matrix <- outer17_species_matrix[,colnames(outer17_species_matrix) %in% outer17_common_species] # filter original species matrix

outer17_rows_to_remove <- rowSums(outer17_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
outer17_species_matrix <- outer17_species_matrix[!outer17_rows_to_remove,]

@@

@@

full_species_matrix <- as.matrix(data[ , -which(names(data) %in% meta_cols)])
full_species_matrix <- full_species_matrix/100 # convert from percent to proportion
rownames(full_species_matrix) <- paste(data$year,"-", data$location, sep = "")

full_species_matrix <- full_species_matrix[rownames(full_species_matrix) %in% relevant_quads,] # subset so only quads that didn't contain rare species are represented
full_species_matrix <- asin(sqrt(full_species_matrix)) # arcsin squareroot transform
data$stratum <- paste(data$year,"-", data$location, sep = "") # for merging with bray matrix later
full_subset <- data[data$stratum %in% relevant_quads,] # make sure rows are also removed from dataframe

brayfull <- vegdist(full_species_matrix, "bray") # create bray dissimilarity matrix
# View(brayfull)
full_braymat <- as.matrix(brayfull) # conver to matrix object

>>>>

bray_mod_site <- glmmTMB(dissimilarity_trans ~ time*habitat + (1 | site),
                              na.action = na.pass, family = beta_family(link = "logit"), data = bray_time_series_site) #
summary(bray_mod_site) 
hist(residuals(bray_mod_site)) # slight skew but not awful
plot(residuals(bray_mod_site) ~ fitted(bray_mod_site)) # some heteroskedasticity but not bad
Anova(bray_mod_site)
emtrends(bray_mod_site, pairwise ~ habitat , "time") # no sig differences
bray_trends_emm_site <- emmip(bray_mod_site, habitat ~ time , at = list(time = seq(1,16,1)), plotit = F, CIs = T)

@@

```{r dissimilarity compared to baseline all, fig.height=15, fig.width = 15}
bray_time_series_site_baseline <- subset(bray_time_series_site, year2 == 2007)
bray_mod_site_baseline <- glmmTMB(dissimilarity_trans ~ time*habitat + (1 | site),
                              na.action = na.pass, family = beta_family(link = "logit"), data = bray_time_series_site_baseline)
hist(residuals(bray_mod_site_baseline)) # long tails but not bad
plot(residuals(bray_mod_site_baseline) ~ fitted(bray_mod_site_baseline) ) # heteroskedastic
Anova(bray_mod_site_baseline)
#                Chisq Df Pr(>Chisq)   
# time          1.9842  1   0.158944   
# habitat      12.3261  3   0.006346 **
# time:habitat  0.7516  3   0.860994   
emtrends(bray_mod_site_baseline, pairwise ~ habitat , "time") # no sig differences
bray_trends_emm_site_baseline <- emmip(bray_mod_site_baseline, habitat ~ time , at = list(time = seq(1,16,1)), plotit = F, CIs = T)

@@

```{r dissimilarity compared to baseline macro, fig.height=15, fig.width = 15}
bray_time_series_site_macro_baseline <- subset(bray_time_series_site_macro, year2 == 2007)
bray_mod_site_macro_baseline <- glmmTMB(dissimilarity_trans ~ time*habitat + (1 | site),
                              na.action = na.pass, family = beta_family(link = "logit"), data = bray_time_series_site_macro_baseline)
hist(residuals(bray_mod_site_macro_baseline)) # slight negative skew
plot(residuals(bray_mod_site_macro_baseline) ~ fitted(bray_mod_site_macro_baseline) ) # strong negative trend
Anova(bray_mod_site_macro_baseline)
Response: dissimilarity_trans
#                Chisq Df Pr(>Chisq)    
# time          0.7391  1    0.38994    
# habitat      56.1360  3  3.929e-12 ***
# time:habitat 14.4254  3    0.00238 ** 
emtrends(bray_mod_site_macro_baseline, pairwise ~ habitat , "time") # fringing slope significantly different from all others
bray_trends_emm_site_macro_baseline <- emmip(bray_mod_site_macro_baseline, habitat ~ time , at = list(time = seq(1,16,1)), plotit = F, CIs = T)

@@

full_columns_to_keep <- apply(full_site_species_matrix, 2, function(x) any(x >= asin(sqrt(0.05) )))
full_abundant_site_species_matrix <- as.data.frame(full_site_species_matrix)[ , full_columns_to_keep]
full_abundant_site_species_names <- colnames(full_abundant_site_species_matrix)

@@

# read in disturbances
disturbances <- read.csv(here("data", "disturbance_list.csv"))
full.cent_site <- merge(full.cent_site, disturbances, by = "year") # add disturbances to path
library(ggrepel)
ggplot() + 
  geom_point(data= full.site.scores_site, aes(x=NMDS1,y=NMDS2, colour = habitat), alpha = 0.05) + # add the point markers
  geom_text_repel(data= full.abundant.species.scores_site,aes(x=NMDS1,y=NMDS2, label = species), size = 6, alpha = 0.7) +  # add the species labels

  geom_path(data = full.cent_site, aes(x = NMDS1, y = NMDS2, group = habitat, colour = habitat),
            arrow = arrow(length=unit(0.4,"cm"), ends="last", type = "closed"), size = 1) +
  geom_point(data = full.cent_site, aes(x = NMDS1, y = NMDS2, group = habitat, colour = habitat, shape = disturbance), size = 3, stroke = 1.4) +
  scale_shape_manual(values = c(
    "cots" = 8,
    "cyclone" = 13,
    "bleaching" = 4
  )) +

@@

full_columns_to_keep <- apply(full_site_macro_species_matrix, 2, function(x) any(x >= asin(sqrt(0.05) )))
full_abundant_site_macro_species_matrix <- as.data.frame(full_site_macro_species_matrix)[ , full_columns_to_keep]
full_abundant_site_macro_species_names <- colnames(full_abundant_site_macro_species_matrix)

@@

----------------- QUADRAT BELOW
```{r quad bray curtis setup}
fringing_species_matrix <- as.matrix(fringing[ , -which(names(fringing) %in% meta_cols)])
fringing_species_matrix <- fringing_species_matrix/fringing$hard_cover # convert from percent to proportion
rownames(fringing_species_matrix) <- paste(fringing$year,"-", fringing$location, sep = "")

# filter rare species
fringing_species_presence <- colSums(fringing_species_matrix > 0) # Presence absence matrix
fringing_species_percentage <- (fringing_species_presence/nrow(fringing_species_matrix)) # proportion of rows where a species is present
fringing_common_species <- names(fringing_species_percentage[fringing_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
# fringing_species_matrix <- fringing_species_matrix[,colnames(fringing_species_matrix) %in% fringing_common_species] # filter original species matrix

fringing_rows_to_remove <- rowSums(fringing_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
fringing_species_matrix <- fringing_species_matrix[!fringing_rows_to_remove,] 
fringing_species_matrix <- na.omit(fringing_species_matrix)

@@

## backreef
backreef_species_matrix <- as.matrix(backreef[ , -which(names(backreef) %in% meta_cols)])
backreef_species_matrix <- backreef_species_matrix/backreef$hard_cover # convert from percent to proportion
rownames(backreef_species_matrix) <- paste(backreef$year,"-", backreef$location, sep = "")

# filter rare species
backreef_species_presence <- colSums(backreef_species_matrix > 0) # Presence absence matrix
backreef_species_percentage <- (backreef_species_presence/nrow(backreef_species_matrix)) # proportion of rows where a species is present
backreef_common_species <- names(backreef_species_percentage[backreef_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
# backreef_species_matrix <- backreef_species_matrix[,colnames(backreef_species_matrix) %in% backreef_common_species] # filter original species matrix

backreef_rows_to_remove <- rowSums(backreef_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
backreef_species_matrix <- backreef_species_matrix[!backreef_rows_to_remove,] 
backreef_species_matrix <- na.omit(backreef_species_matrix)

@@

## Outer 10
outer10_species_matrix <- as.matrix(outer10[ , -which(names(outer10) %in% meta_cols)])
outer10_species_matrix <- outer10_species_matrix/outer10$hard_cover # convert from percent to proportion
rownames(outer10_species_matrix) <- paste(outer10$year,"-", outer10$location, sep = "")

# filter rare species
outer10_species_presence <- colSums(outer10_species_matrix > 0) # Presence absence matrix
outer10_species_percentage <- (outer10_species_presence/nrow(outer10_species_matrix)) # proportion of rows where a species is present
outer10_common_species <- names(outer10_species_percentage[outer10_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
# outer10_species_matrix <- outer10_species_matrix[,colnames(outer10_species_matrix) %in% outer10_common_species] # filter original species matrix

outer10_rows_to_remove <- rowSums(outer10_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
outer10_species_matrix <- outer10_species_matrix[!outer10_rows_to_remove,] 
outer10_species_matrix <- na.omit(outer10_species_matrix)

@@

## OUTER 17
outer17_species_matrix <- as.matrix(outer17[ , -which(names(outer17) %in% meta_cols)])
outer17_species_matrix <- outer17_species_matrix/outer17$hard_cover # convert from percent to proportion
rownames(outer17_species_matrix) <- paste(outer17$year,"-", outer17$location, sep = "")

# filter rare species
outer17_species_presence <- colSums(outer17_species_matrix > 0) # Presence absence matrix
outer17_species_percentage <- (outer17_species_presence/nrow(outer17_species_matrix)) # proportion of rows where a species is present
outer17_common_species <- names(outer17_species_percentage[outer17_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
# outer17_species_matrix <- outer17_species_matrix[,colnames(outer17_species_matrix) %in% outer17_common_species] # filter original species matrix

outer17_rows_to_remove <- rowSums(outer17_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
outer17_species_matrix <- outer17_species_matrix[!outer17_rows_to_remove,] 
outer17_species_matrix <- na.omit(outer17_species_matrix)

@@

```{r quad level beta dispersion}
full_species_matrix <- as.matrix(data[ , -which(names(data) %in% meta_cols)])
full_species_matrix <- full_species_matrix/data$hard_cover # convert from percent to proportion
rownames(full_species_matrix) <- paste(data$year,"-", data$location, sep = "")
full_rows_to_remove <- rowSums(full_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
full_species_matrix <- full_species_matrix[!full_rows_to_remove,] 
full_species_matrix <- na.omit(full_species_matrix)
full_species_matrix <- asin(sqrt(full_species_matrix)) # arcsin squareroot transform
data$stratum <- paste(data$year,"-", data$location, sep = "") # for merging with bray matrix later
full_subset <- data[data$stratum %in% rownames(full_species_matrix),] # make sure rows are also removed from dataframe

brayfull <- vegdist(full_species_matrix, "bray") # create bray dissimilarity matrix
# full_quad_betadisper <- vegan::betadisper(brayfull, full_subset$hab_year)
full_subset$betadispersion <- full_quad_betadisper$distances[match(full_subset$stratum, names(full_quad_betadisper$distances))]

full_subset$betadispersion_trans <- sv_trans(full_subset$betadispersion, full_subset)

betadisper_quad_mod <- glmmTMB(betadispersion_trans ~ year*habitat + (1|site), family = beta_family(link = "logit"), data = full_subset)
summary(betadisper_quad_mod) # fair bit of within site variance - inv_logit(0.02) = 0.5
hist(residuals(betadisper_quad_mod)) # positive skew
plot(residuals(betadisper_quad_mod) ~ fitted(betadisper_quad_mod)) # slight heteroskedasticity
Anova(betadisper_quad_mod)
#                Chisq Df Pr(>Chisq)    
# year           0.3156  1     0.5743    
# habitat      333.7035  3     <2e-16 ***
# year:habitat 720.3054  3     <2e-16 ***

macro_species_matrix <- as.matrix(macroalgae[ , -which(names(macroalgae) %in% meta_cols)])
macro_species_matrix <- macro_species_matrix/macroalgae$hard_cover # convert from percent to proportion
rownames(macro_species_matrix) <- paste(macroalgae$year,"-", macroalgae$location, sep = "")
macro_rows_to_remove <- rowSums(macro_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
macro_species_matrix <- macro_species_matrix[!macro_rows_to_remove,] 
macro_species_matrix <- na.omit(macro_species_matrix)
macro_species_matrix <- asin(sqrt(macro_species_matrix)) # arcsin squareroot transform
macroalgae$stratum <- paste(macroalgae$year,"-", macroalgae$location, sep = "") # for merging with bray matrix later
macro_subset <- macroalgae[macroalgae$stratum %in% rownames(macro_species_matrix),] # make sure rows are also removed from macroalgaeframe

braymacro <- vegdist(macro_species_matrix, "bray") # create bray dissimilarity matrix
# macro_quad_betadisper <- vegan::betadisper(braymacro, macro_subset$hab_year)
macro_subset$betadispersion <- macro_quad_betadisper$distances[match(macro_subset$stratum, names(macro_quad_betadisper$distances))]

@@

full_braymat <- as.matrix(brayfull) # conver to matrix object
